Brownian motion of main-belt asteroids on human timescales

### 2025-02-18 main.tex
sigmaM edits.

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                              \frac{dn}{dH\,dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int db_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{\sigma_M(H)}{bv}\right)^2 \\
  & = & \int dH \frac{dN}{dH} \sigma_M(H)^2  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        2\pi \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be determined from fitting the data of the tracer.
Fitting an ephemeris that includes all of these Individually
detectable encounters is entirely feasible, as long as the relevant
MBAs are known, with modestly accurate orbits.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$

@@

\subsubsection{Encounter rates and geometries}
The pairwise interaction rate $dn/dA\,dv$ and the impulse direction
distributions $\langle \hat b^2_\gamma \rangle$ are estimated from a
numerical integration of all of the MBA orbits available
from the Minor Planet Center (MPC) as of ???? 2025.   We retain as
potential deflectors the 1.26~million bodies with $1.8<a<4.2$~AU,
observations at multiple oppositions, and uncertainty values $U\le5.$
We designate each source as being an ``Inner'' MBA, with semi-major axes between the 4:1 and 3:1
mean-motion resonances of Jupiter; ``Middle'' MBA between the 3:1 and
5:2 resonances; ``Outer'' MBAs between the 5:2 and 2:1
resonances; and the small remainder as ``Other.''  Our working
assumption will be that the MPC objects are an unbiased sample of the
orbital-element distribution within each of the Inner, Middle, and
Outer belt regions.

From this full MBA list, we select a random subset of 5000 objects
from each of the Inner, Middle, and Outer belts to serve as a sample
of ``tracers.''  We record all of the passages of any tracer MBA
within 0.03~AU of any other ``deflector'' MBA (drawn from the full
catalog of 1.26~million) over a 10-year period.  The 5000 tracers per
belt region are enough to be representative of the statistics of
the region, but few enough that the calculation can be done easily
on a laptop computer.
We will report $\langle nI_\gamma^2\rangle$
separately for tracers of each region, allowing for the fact that a
tracer of a given region can receive impulses from deflectors resident
in all regions.

Using a simple leapfrog integrator with gravity from the Sun and the 8
major planets, we advance all MBA's orbits from their heliocentric
osculating elements at the epochs given by the MPC, to barycentric
state vectors on 1 May 2025.  We then find all tracer-deflector pairs
that approach within 0.03~AU of each during the $\pm1$-day period
around this initial epoch, assuming inertial relative motion during
this interval.  The circumstances of such encounters are saved: the
identities of the two MBAs involved, the time of closest approach, the
relative velocity $v,$ and the impact parameter vector $\vecb.$

The leapfrog integrator advances all 1.26 million state vectors by 2
days, and the process is repeated.  We continue at 2-day intervals
until we have recorded 10 years' worth of encounters---17.7 million
total impulses.

@@

For each pair of tracer-deflector regions (\eg Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$.  Figure~\ref{fig:rtz} plots these functions of $v$
for the Inner-Inner, Middle-Middle, and Outer-Outer scattering
events.  Here it is apparent that the $\hat b_\phi$ component is
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.

@@

The two largest asteroids, Ceres and Vesta, have masses determined to
high precision by the \textit{Dawn} spacecraft.  Thus despite holding
most of the mass of the MBAs, their $\sigma_M$ values are just
$\approx3\times10^{-15} M_\odot$ and
the unknown part of their contributions to other MBA's motion are
quite small.  Aside from a handful of other asteroids in
well-characterized binary systems or visited by spacecraft, most other knowledge of MBA masses
comes from mutual gravitational encounters.
\citet{negin} forecast the extent to which LSST data will be able to
constrain the masses of individual MBAs by measuring deflections of
the $\approx 10$ tracer asteroids they encounter most closely over the
course of the nominal 10-year survey.  They find that the $1\sigma$
statistical error on a typical MBA is $\sigma_M=???\, M_\odot.$

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{mbacounts.pdf}
  \caption{Differential counts of MBAs cataloged by the MPC are
    plotted vs absolute magnitude $H$  
    for the Inner, Middle, Outer, and ``Other'' regions of
    the main belt, and for their total.  We assume that they are
    complete for $H\le17,$ and extrapolate to fainter sources using
    the functional form from \eqq{eq:lsstmba} shown as the dotted
    curve.}
  \label{fig:counts}
\end{figure}

Figure~\ref{fig:counts} plots our assumptions for $dN/dH.$  The MPC
catalog is likely close to complete in all three regions for $H\le17,$
so we will use the MPC counts directly in this regime.  Note that the
three regions have similar shapes of $dN/dH.$  For $H>17,$ we adopt
the functional form for the cumulative MBA counts given by
initial LSST predictions \citep{LSST_science_book}:
\begin{equation}
  N(<H) \propto \frac{10^{0.43(H-15.7)}}{10^{0.18(H-15.7)} +
    10^{-0.18(H-15.7)}},
\label{eq:lsstmba}
\end{equation}
normalizing this curve to the observed $H<17$ counts for each region.

\subsubsection{Individually detectable impulses}
For several \citet{negin} forecast the uncertainty $\sigma_v$ on the impulse 
applied to each $\approx7000$ tracers using astrometric data expected from LSST
over its 10-year duration.  The value of $\sigma_I$ depends upon the
apparent magnitude of the tracer, the timing and geometry of the
impulse, and the stochastic variations in the observing pattern and
weather for LSST.
For impulses occurring near the midpoint
of the survey, $\approx80\%$ of all cases yield $\sigma_I<10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ some an order of magnitude lower.  We conservatively
assume that $I_{\rm det}=5\times 10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ which when normalized by $v_c$ yields $I_{\rm
  det}=10^{-8}$ at 10\% accuracy across the main belt.

@@

>>>>

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                              \frac{dn}{dH\,dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int db_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{\sigma_M(H)}{bv}\right)^2 \\
  & = & \int dH \frac{dN}{dH} \sigma_M(H)^2  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        2\pi \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be determined from fitting the data of the tracer.
Fitting an ephemeris that includes all of these Individually
detectable encounters is entirely feasible, as long as the relevant
MBAs are known, with modestly accurate orbits.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$

@@

\subsubsection{Encounter rates and geometries}
The pairwise interaction rate $dn/dA\,dv$ and the impulse direction
distributions $\langle \hat b^2_\gamma \rangle$ are estimated from a
numerical integration of all of the MBA orbits available
from the Minor Planet Center (MPC) as of 6 January 2025.   We retain as
potential deflectors the 1.26~million bodies with $1.8<a<4.2$~AU,
observations at multiple oppositions, and uncertainty values $U\le5.$
We designate each source as being an ``Inner'' MBA, with semi-major axes between the 4:1 and 3:1
mean-motion resonances of Jupiter; ``Middle'' MBA between the 3:1 and
5:2 resonances; ``Outer'' MBAs between the 5:2 and 2:1
resonances; the remaining ``Other'' objects are a small minority that
contribute very little to the total diffusion rate, and we will ignore.  Our working
assumption will be that the MPC objects are an unbiased sample of the
orbital-element distribution within each of the Inner, Middle, and
Outer belt regions.

From this full MBA list, we select a random subset of 5000 objects
from each of the Inner, Middle, and Outer belts to serve as a sample
of ``tracers.''  We record all of the passages of any tracer MBA
within 0.03~AU of any other ``deflector'' MBA (drawn from the full
catalog of 1.26~million) over a 10-year period.  The 5000 tracers per
belt region are enough to be representative of the statistics of
the region, but few enough that the calculation can be done easily
on a laptop computer.
We will estimate the constituents of  $\langle nI_\gamma^2\rangle$ in
\eqq{eq:nI2} separately for each of the nine combinations of tracers
in Inner, Middle, and Outer regions with deflectors from each region.

Using a simple leapfrog integrator with gravity from the Sun and the 8
major planets, we advance all MBA's orbits from their heliocentric
osculating elements at the epochs given by the MPC, to barycentric
state vectors on 1 May 2025.  Using $kd$-trees to accelerate
pair-finding, we locate all tracer-deflector pairs
that approach within 0.03~AU of each during the $\pm1$-day period
around this initial epoch, assuming inertial relative motion during
this interval.  The circumstances of such encounters are saved: the
identities of the two MBAs involved, the time of closest approach, the
relative velocity $v,$ and the impact parameter vector $\vecb.$

The leapfrog integrator advances all 1.26 million state vectors by 2
days, and the process is repeated.  We continue searching for pairs at
2-day intervals until we have recorded 10 years' worth of encounters---17.7 million
total impulses.

@@

For each pair of tracer-deflector regions (\eg\ Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$.  Figure~\ref{fig:rtz} plots these functions of $v$
for the Inner-Inner, Middle-Middle, and Outer-Outer scattering
events.  Here it is apparent that the $\hat b_\phi$ component is
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.  An important result is that
the quantity $v^{-2} (dN/dA\,dv)$ is always well-behaved as $v\rightarrow 0,$
so we do not need any special treatment to sum the integrals in
\eqq{eq:nI2}.  The $v$ integrand peak is in the range $0.03<v/v_c<0.1.$

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{mbacounts.pdf}
  \caption{Differential counts of MBAs cataloged by the MPC are
    plotted vs absolute magnitude $H$  
    for the Inner, Middle, Outer, and ``Other'' regions of
    the main belt, and for their total.  We assume that they are
    complete for $H\le17,$ and extrapolate to fainter sources using
    the functional form from \eqq{eq:lsstmba} shown as the dotted
    curve.}
  \label{fig:counts}
\end{figure}
We also require the $H$ distribution of potential deflectors.
Figure~\ref{fig:counts} plots our assumptions for $dN/dH.$  The MPC
catalog is likely close to complete in all three regions for $H\le17,$
so we will use the MPC counts directly in this regime.  Note that the
three regions have similar shapes of $dN/dH.$  For $H>17,$ we adopt
the functional form for the cumulative MBA counts given by
initial LSST predictions \citep{LSST_science_book}:
\begin{equation}
  N(<H) \propto \frac{10^{0.43(H-15.7)}}{10^{0.18(H-15.7)} +
    10^{-0.18(H-15.7)}},
\label{eq:lsstmba}
\end{equation}
normalizing this curve to the observed $H<17$ counts for each region.

Our integral for the uncertainty due to unmodelled MBA gravitation
requires the RMS error $\sigma_M(H)$ between the true mass $M$ and the
mass used (if any) in the ephemeris model.
The two largest asteroids, Ceres and Vesta, have masses determined to
high precision by the \textit{Dawn} spacecraft.  Thus despite holding
half of the mass of the asteroid belt, their $\sigma_M$ values are just
$\approx3\times10^{-15} M_\odot$ and
the unknown part of their contributions to other MBA's motion are
quite small.  Aside from a handful of other asteroids in
well-characterized binary systems or visited by spacecraft, most other knowledge of MBA masses
has and will come from mutual gravitational encounters.   An
individual asteroid's mass can be estimated from its measured effect  on the
few MBAs on which it imparts the largest impulses.  The accuracy
$\sigma_M$ of this determination will depend upon the geometry of
these encounters, the measurement errors on the tracers, and the size
of the noise from unmodelled mass that we aim to determine.  These
factors are independent of the mass of the deflector.  We will
assign a value $\sigma_M$ to the RMS uncertainty in MBA mass
attainable from following a few individual tracers, and calculate
the Brownian-motion noise as a function of $\sigma_M.$  We take
\begin{equation}
  \sigma_M(H) = \textrm{min}\left[ \sigma_M, M(H) \right],
\end{equation}
because a given MBA could have a mass determination accurate to
$\sigma_M$ if included in the ephemeris model, or if $M(H)<\sigma_M,$
it would be omitted from the ephemeris model.

To give some context, the most recent
comprehensive investigation of MBA masses from mutual scattering
events was by \citet{baer}, who report typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs.  We can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry.

\subsubsection{Individually detectable impulses}
\citet{negin} forecast the uncertainty $\sigma_v$ on an impulse 
applied to each of $\approx7000$ tracers using astrometric data expected from LSST
over its 10-year duration.  The value of $\sigma_I$ depends upon the
apparent magnitude of the tracer, the timing and geometry of the
impulse, and the stochastic variations in the observing pattern and
weather for LSST.
For impulses occurring near the midpoint
of the survey, $\approx80\%$ of all cases yield $\sigma_I<10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ some an order of magnitude lower.  We conservatively
assume that $I_{\rm det}=5\times 10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ which when normalized by $v_c$ yields $I_{\rm
  det}=10^{-8}$ at 10\% accuracy across the main belt.

@@

\citet{negin} forecast the extent to which LSST data will be able to
constrain the masses of individual MBAs by measuring deflections of
the $\approx 10$ tracer asteroids they encounter most closely over the
course of the nominal 10-year survey.  They find that the $1\sigma$
statistical error on a typical MBA is $\sigma_M=???\, M_\odot.$
